Characterizing dynamical transitions by statistical complexity measures based on ordinal pattern transition networks

### 2020-11-23 manuscript_Huang.tex
Updates from Overleaf

\begin{abstract}
In the recent decade, complex network approaches have been emerging as novel and complementary concepts of nonlinear time series analysis that are able to unveil many features that are hidden to more traditional analysis methods. In this work, we focus on one particular approach: the application of ordinal pattern transition network for characterizing time series data. More specifically, we generalize some traditional estimators of statistical complexity measures (SCM), which explicitly disclose heterogeneous frequencies of ordinal pattern transitions in the time series networks. To demonstrate the usefulness of the estimators, we employ SCM to characterize different dynamical transitions in the paradigmatic model system of the logistic map as well as real-world time series of fluid experiments and ECG data. The obtained results for both, artificial and experimental data demonstrate that the consideration of transition frequencies between different ordinal patterns improves the estimation of SCM, which is especially important for the analysis of time series with finite length. 
\end{abstract}

@@

\begin{quotation}
In the recent decade, tools of traditional nonlinear time series analysis have been undergoing fast developments benefiting concepts from complex network theories. Along this line of research topics, ordinal pattern transition time series network has been providing complement insights to the permutation entropy estimators for time series data. Traditionally, permutations based on order patterns are simple and easily implemented tools for estimators of statistical complexity measures (SCMs), which however rely on the static pattern frequencies only. Still, much additional knowledge of the order pattern transitions to the SCMs has not been fully considered. In this work, we generalize the estimators of SCMs by means of ordinal pattern transition networks which take into account the pattern transition properties explicitly. The usefulness of the generalizations will be demonstrated by time series of both model and experimental data. 
\end{quotation}

\section{Introduction}
In the recent decade, complex network based approaches have emerged as prominent tools for nonlinear time series analysis \cite{Kantz97,Sprott2003}, which have found many interesting applications to experimental data \cite{ZouPR2018}. In this line of research, there are several major methods are of particular importance, including recurrence network \cite{MarwanPLA2009,Donner2010a}, visibility graphs \cite{Lacasa2008,Nunez2012}, transition networks \cite{Nicolis2005,MichaelChaos2015} and cycle network \cite{ZhangPRL2006}. Depending on the particular questions at hand, we note that each of these complex network approaches has various variants for different applications, for instance, constructing a single network for a univariate time series or interacting networks (multiplex or multilayer) networks for coupled time series, which has been systematically reviewed in \cite{ZouPR2018} recently. In this work, we focus on constructing ordinal pattern transition networks for time time series data, which show the advantages of easy implementations and wide applications to various data from different origins, including correlated stochastic processes, EEG and human cardiac series (ECG) \cite{MichaelChaos2015,KulpChaos2016,zhangSciRep2017,McCullough2017b,BorgesAMC2019}.

The basic idea of ordinal partition network method originates from identifying ordinal patterns of time series \cite{BandtPRL2002}, which is a well developed concept in nonlinear time series analysis. For instance, permutation entropy and the associated statistical analysis \cite{BandtPRL2002}. In addition, ordinal symbolic representation of time series has found a number of interesting applications in science and engineering, for instance, biomedical recordings \cite{AmigoPTRSA2014}, finance \cite{ZaninChaos2008}, climate sciences \cite{BarreiroChaos2011}. Some recent progress has been comprehensively reviewed in Ref. \cite{AmigoPTRSA2014}. More specifically, given a one-dimensional time series $\{ x(i)\}_{i=1, \cdots, L}$ comprising of $L$ sampled points from a dynamical system, we first reconstruct the corresponding phase space by time delay embedding $\vec{x}(i) = [x(i), x(i+\tau), \cdots, x(i+(D-1)\tau)]$ with dimension $D$ \cite{Takens1981,Kantz97}. Next, we represent each embedded phase point $\vec{x}(i)$ by its corresponding rank order, which is represented by a symbol $\pi_{i}$. When sliding windows from $i=1$ to $N = L - (D - 1)\tau$ in the embedded space, a symbolic representation of the trajectory $\pi_i$ is produced. It is possible to derive information about the dynamics of the underlying system by assessing different probabilities of ordinal patterns which is typical for a time series resulted from a deterministic system. More generally, we obtain a probability distribution $P$ whose elements $p(\pi_{i})$ are the frequencies associated to pattern $\pi_{i}$, $i = 1, \cdots, D!$. Clearly, $P$ provides a significant feasible way to estimate the probability distribution function for a given time series, which plays a crucial role in extracting statistical complexities of the underlying dynamical systems \cite{BandtPRL2002,AmigoPTRSA2014}.

Relatively easy computation of $P$ by ordinal patterns for the underlying system prompts another successful application to compute statistical complexity measure \cite{kowalskiEntropy2011}. Traditionally, statistical complexity measures (SCMs) are defined as the product of the normalized Shannon entropy and disequilibrium \cite{kowalskiEntropy2011}, which capture specific organizational properties of structure and patterns in a process. Following the literature \cite{rossoPRL2007,LopezPLA1995,kowalskiEntropy2011}, in this work we consider Jensen-Shannon SCM which is defined as $\mathcal{C}_{JS}[P] = \mathcal{H}[P]  \cdot  Q_{JS}[P, P_e]$, where $\mathcal{H}[P]$ is the normalized Shannon entropy and $Q_{JS}[P, P_e]$ is the so-called disequilibrium function measuring the distance of the given $P$ to the uniform distribution $P_e$, i.e., by the Jensen-Shannon divergence. Thus, $Q_{JS}[P, P_e]$ is defined as $Q_{JS}[P, P_e] = Q_{0} \cdot JS[P, P_e]$ where $JS[P, P_e]$ is the Jensen-Shannon divergence between $P$ and $P_e$ and $Q_{0}$ is the normalization factor. Furthermore, the complexity-entropy plane $\mathcal{H} \times \mathcal{C}_{JS}$ has been intensively used to distinguish chaotic systems from stochastic ones \cite{rossoPRL2007}. The computation of the complexity-entropy planes has benefit from the ordinal pattern symbolization of time series and has found various applications \cite{rossoPRL2007,kowalskiEntropy2011}. However, in the complexity-entropy analysis, there are some chaotic maps that could be easily confused as random noise because no clear separable regions are available to differentiate dynamics as reported in \cite{BorgesAMC2019}. So far, it remains to be a challenging task to disclose the possible non-monotonic relationship between complexity and entropy \cite{MartinPLA2003}.

Note that, in all above discussions of ordinal pattern analysis, the transition behavior between patterns remains largely unconsidered. Recently, ordinal partition network representations have been proposed to capture the transition behavior of the ordinal patterns \cite{MichaelChaos2015,KulpChaos2016,zhangSciRep2017,McCullough2017b,BorgesAMC2019}, which sheds novel insight on the standard ordinal symbolic analysis of time series. In particular, each ordinal pattern is considered as a vertex in the graph and a directed weighted edge connecting two patterns in the graph is established depending on the transition frequency. One of advantages of ordinal pattern transition networks is that we can provide pronounced distinction among different time series with short time series \cite{MichaelChaos2015,BorgesAMC2019}.

In this work, we generalize the traditional statistical complexity measures by incorporating pattern transition probabilities. The advantages are the following: (i) All SCMs successfully show distinctions between different dynamics, including time series from numerical models and experimental data; (ii) These generalized SCMs help to capture the consistent relationship between statistical complexity measures and chaoticity; (iii) All SCMs are sensitive to dynamical transitions of period doubling, band merging, inner and outer crisis.

The structure of the paper is organized as follows: In section \ref{sec:OPW}, we show two slightly different ways to obtain the transition matrix of the corresponding ordinal pattern transition networks. Furthermore, based on this matrix, we propose in section \ref{sec:SCM} to compute the statistical complexity measures from three different perspectives including static pattern frequency properties, dynamic pattern transition properties and both. We will use one example to emphasize the important effects of embedding parameters in section \ref{sec:embeddings}. The usefulness of SCMs will be illustrated by four time series of typical dynamical regimes of the Logistic map. In section \ref{sec:plane}, the traditional complexity-entropy planes of each SCM will be discussed while the control parameter is systematically changed. In order to characterize the dynamical transitions, we expand the SCM values versus the control parameter in section \ref{sec:transi}. Finally, we show that SCMs successfully distinguish different dynamical regimes by experimental time series from flow data and human ECG data in \ref{sec:time}. Some discussions will be provided in the last section \ref{sec:con}.

\section{Ordinal pattern transition matrix} \label{sec:OPW}
Most of the current studies compute permutation entropy $\mathcal{S}_O$ focusing on the frequencies of ordinal patterns, which however overlooked the transition behavior between  patterns. To emphasize the transitions between any pair of two patterns, measures of transitional complexity have been further proposed in \cite{zhangSciRep2017,McCullough2017b} to quantify the resulting OPTNs. To this end, we indicate each directed link by its transition frequency $w_{ij} = p({\pi_i \to \pi_j})$, following the time iterations of the series. Calculating the transition frequency for each pattern, we obtain a weighted directed network characterized by a weighted adjacency matrix $W = \{ w_{ij} \}, i, j \in [1, D!]$.

Based on the transition matrix $\mathbf{W}$ of an OPTN, we note that there two slightly different ways to introduce normalization factors in $W$. The first option is to use a global normalization to ensure $\sum_{i,j}^{D!} w_{ij} = 1$ \cite{zhangSciRep2017}, which results in a globally normalized transition entropy as termed below.

The second option of normalization has been proposed by McCullough {\textit{et al}} \cite{McCullough2017b} that they first considered the local out-link transition frequency from pattern $\pi_i$ to $\pi_j$
\begin{equation} \label{eq:localTp}
p_{\pi_i \to \pi_j} = \left \{ \begin{aligned}
				& 0,  \;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\; \text{if} \;\;\; \pi_i = \pi_j \\
				& \frac{w_{ij}}{\sum_{j, j \neq i} w_{ij}}, \;\;\;\;\;\; \text{if} \; \pi_i \neq \pi_j.\\
				\end{aligned}
				\right.
\end{equation}		
Note that the transition frequency of Eq.~\eqref{eq:localTp} is pattern (row-) wise normalized. This case will be referred to as node-wise out-link normalized transition matrix $\mathbf{W}$ in the following. In this way, it easily captures heterogeneous behavior of both static frequencies of different patterns and dynamic transition frequencies between patterns.

In addition, we need to consider self-loops that may affect the numerical estimation of $\mathbf{W}$ which is especially important for time series from continuous systems. More specifically, it has been demonstrated in \cite{zhangSciRep2017} that there are about $99\%$ self-loops in the time series of R\"ossler system integrated by a step size $h = 0.01$, while about $1\%$ non-self-loops are hidden in $\mathbf{W}$. In addition, self-loops are related to the temporal correlation of time series \cite{BorgesAMC2019}. On the other hand, neglecting self-loops in $\mathbf{W}$ emphasizes the transition behavior between different patterns, which follows most of the existing studies of complex networks for computational simplicity and theoretical concerns \cite{CostaADPhy2007}. Therefore, we remove self-loops before computing the weighted matrix $\mathbf{W}$. In other cases of stochastic processes and discrete systems, self-loops have been included in the analysis showing some interesting results $\mathbf{W}$ \cite{BorgesAMC2019}. We argue that whether to remove or to consider self-loops depends on the particular processes under study. Anyway, we have verified that our results below do not change significantly while considering effects of self-loops.

\section{Statistical complexity measures} \label{sec:SCM}
In this section, we review briefly ordinal pattern frequency based SCM and propose two novel ways to define SCM, which explicitly consider transition frequencies of the resulting ordinal network representations for time series.

$\bullet$ SCM based on permutation entropy of pattern frequencies.

The permutation entropy of a time series is the Shannon entropy of the distribution of patterns $P = \{p_{\pi_1}, p_{\pi_2}, \cdots, p_{\pi_i} \}$, which is computed as
\begin{equation}
\mathcal{S}_{O}[P]= - \sum_{i=1}^{D!} p_{\pi_i} \log p_{\pi_i}. 
\end{equation}
The normalized permutation entropy $\mathcal{H}_O$ is defined as 
\begin{equation} \label{eq:Ho}
\mathcal{H}_{O}[P] = \mathcal{S}_{O}[P] / \mathcal{S}_{O, max}, 
\end{equation}
where $\mathcal{S}_{O, max} = \log D!$ is resulted from the case of uniform distribution of the same number of patterns $P_e = \{1/D!, \cdots, 1/D!\}$.  In the next step, we use some distance definition to measure the given $P$ to the uniform distribution $P_e$, which is the core of statistical complexity measures (SCM).  There are several different distance functions as illustrated in the literature, such as Euclidean norm, Wooters' distance, Kullback-Leibler relative entropy and Jensen-Shannon divergence \cite{kowalskiEntropy2011}. In this work, the difference between the distribution $P$ and $P_e$ is characterized by the Jensen-Shannon divergence $Q_{JS}[P] = Q_0 \cdot JS[P, P_e]$ where $JS[P, P_e]$ is calculated by 
\begin{equation}
JS[P, P_e] = \mathcal{S}_O[\frac{P+P_e}{2}] - \frac{1}{2}\mathcal{S}_O[P] - \frac{1}{2}\mathcal{S}_O[P_e], 
\end{equation}
where $Q_0$ is a normalization factor with its value equal to the inverse of the maximal distance $JS[P, P_e]$, which ensures $Q_{JS}[P] \in [0, 1]$. This maximal distance is obtained when the distribution $P$ has probability one at one pattern and zeros at all other patterns. The general expression of $Q_0$ reads  
\begin{equation} \label{eq:Q0}
Q_0 = -2 \{(\frac{N+1}{N}) \log (N+1) -2 \log (2N) + \log (N)  \}^{-1}
\end{equation}
where $N = D!$ is the number of maximal patterns. For large $D$, we have $Q_0 \approx 1 /  \log 2$.

The function $Q_{JS}[P]$ is different from zero if the system has a kind of privilege over patterns among the accessible ones, reflecting determinism of the system. Finally, the SCM based on permutation entropy of node (pattern) frequencies is defined as 
\begin{equation}
\mathcal{C}_{O}[P] = Q_{JS}[P] \cdot \mathcal{H}_{O}[P].
\end{equation}
This quantity is static in the sense that it quantifies the amount of information stored by ordinal frequencies in the system and its disequilibrium of the observed parts of its accessible states in comparison to uniform distribution \cite{LopezPLA1995}.

$\bullet$ SCM based on transition entropy by globally normalized weighting matrix $W$.

In this case, the sum of all entries of the transition matrix $W$ is 1 and the Shannon entropy of transition frequencies between any pair of ordinal patterns is therefore 
\begin{equation}
\begin{split}
\mathcal{S}_{W}[W] &= - \sum_{i=1}^{D!} \sum_{j = 1; j \neq i}^{D!} w_{ij} \log w_{ij}, \\
& = - \sum_{i=1}^{D!} \sum_{j =1; j \neq i}^{D!} p_{\pi_i \to \pi_j} \log p_{\pi_i \to \pi_j}. 
\end{split}
\end{equation}
The normalized entropy is hence defined 
\begin{equation}
\mathcal{H}_{W}[W] = \mathcal{S}_{W}[W] / \mathcal{S}_{W, max}, 
\end{equation}
where $\mathcal{S}_{W, max} = \log D! (D! - 1)$ and the self transitions are excluded. Furthermore, there are $D! (D! - 1)$ pattern transitions in the uniformly distribution function $P_e$. Thus, in a full analogy with node-based SCM, the Jensen-Shannon distance is therefore computed as $Q_{JS}[W] = Q_0 \cdot JS[W, P_e]$, where $JS[W, P_e]$ is computed as 
\begin{equation}
JS [W, P_e] = \mathcal{S}_{W}[\frac{W + P_e}{2}] - \frac{1}{2}\mathcal{S}_{W}[W] - \frac{1}{2}\mathcal{S}_{W}[P_e]. 
\end{equation}
We note that we have the same expression of the normalization factor $Q_0$ while replacing $N = D! (D! - 1)$ in Eq. \eqref{eq:Q0}. This further yields the SCM 
\begin{equation}
\mathcal{C}_{W}[W] = Q_{JS}[W] \cdot \mathcal{H}_{W}[W].
\end{equation}
This quantify is dynamic in the sense that it considers the state transition information stored in the system.

$\bullet$ SCM based on transition entropy by node-wise out-edge normalized
weighting matrix $W$.

>>>>

\begin{abstract}
Complex network approaches have been recently emerging as novel and complementary concepts of nonlinear time series analysis which are able to unveil many features that are hidden to more traditional analysis methods. In this work, we focus on one particular approach: the application of ordinal pattern transition networks for characterizing time series data. More specifically, we generalize some traditional estimators of statistical complexity measures (SCMs), which explicitly disclose heterogeneous frequencies of ordinal pattern transitions in the time series networks. To demonstrate the usefulness of the estimators, we employ SCMs to characterize different dynamical transitions in the logistic map as a paradigmatic model system, as well as real-world time series of fluid experiments and ECG data. The obtained results for both, artificial and experimental data demonstrate that the consideration of transition frequencies between different ordinal patterns leads to dynamically meaningful estimates of SCMs, which provide prospective tools for the analysis of time series with finite length. 
\end{abstract}

@@

\begin{quotation}
In the recent decade, traditional tools of nonlinear time series analysis have been undergoing fast developments benefiting from concepts from complex network theory. Along this line of research topics, ordinal pattern transition networks have been expanding the established concept of permutation entropy and provide complementary insights into the dynamical organization underlying time series data. Traditionally, permutations based on order patterns are simple and easy to implement concepts providing estimators of statistical complexity measures (SCMs), which however rely on static pattern frequencies only. Yet, much additional information can be exploited by including order pattern transitions into the definitions of SCMs -- an idea that however has not been widely developed and applied so far. In this work, we generalize existing estimators of SCMs by means of ordinal pattern transition networks which take into account the pattern transition properties explicitly. The usefulness of our generalizations of existing SCMs is demonstrated by using time series of both, model and experimental data. 
\end{quotation}

\section{Introduction}
In the recent decade, complex network based approaches have emerged as prominent tools for nonlinear time series analysis \cite{Kantz97,Sprott2003}, which have found many interesting applications to experimental data from various origins \cite{ZouPR2018}. In this line of research, there exist several major methods that are of particular importance, including recurrence networks \cite{MarwanPLA2009,Donner2010a}, visibility graphs \cite{Lacasa2008,Nunez2012}, transition networks \cite{Nicolis2005,MichaelChaos2015} and cycle networks \cite{ZhangPRL2006}. Depending on the particular questions at hand, each of these complex network approaches exhibit various variants that can be employed for different application purposes. For instance, we may either construct a single network representing a univariate time series, or interacting, multiplex or multi-layer networks for coupled time series. For a recent systematic review, see Ref.~\cite{ZouPR2018}. In this work, we focus on constructing ordinal pattern transition networks (OPTNs) for time series data, which present the advantages of easy implementation and a wide range of existing application to data from different origins, including correlated stochastic processes, neurophysiological (EEG) and human cardiac activity (ECG) series \cite{MichaelChaos2015,KulpChaos2016,zhangSciRep2017,McCullough2017b,BorgesAMC2019,Subramaniyam2020}.

The basic idea behind the OPTN method originates from identifying ordinal patterns in time series \cite{BandtPRL2002}, which is a well developed concept in nonlinear time series analysis leading to, for instance, the widely employed measure of permutation entropy \cite{BandtPRL2002}. In addition, ordinal symbolic representations of time series have found a number of interesting applications in science and engineering, for instance, in biomedical recordings \cite{AmigoPTRSA2014}, finance \cite{ZaninChaos2008}, and even climate \cite{BarreiroChaos2011}. Some recent progress has been comprehensively reviewed in Ref.~\cite{AmigoPTRSA2014}.

More specifically, given a one-dimensional time series $\{ x_i\}_{i=1, \cdots, L}$ comprising $L$ observations from a dynamical system, we first reconstruct the corresponding phase space by time delay embedding $\vec{x}_i = [x_i, x_{i+\tau}, \ldots, x_{i+(D-1)\tau}]$ with an embedding dimension $D$ \cite{Takens1981,Kantz97}. Next, we represent each embedding vector $\vec{x}(i)$ by the corresponding rank order of its components, which is encoded into a symbol $\pi_{i}$. Hence, when sliding windows from $i=1$ to $N = L - (D - 1)\tau$ in the embedding space, a symbolic representation of the trajectory $\{\pi_i\}$ is produced. It is possible to derive information about the dynamics of the underlying system by assessing the probabilities of occurrence of different ordinal patterns. For example, for time series resulting from deterministic dynamical systems, certain patterns may not occur at all. More generally, we obtain an empirical probability distribution $P$ whose elements $p_{\pi_{i}}$ are the frequencies associated to the different patterns $\pi_{i}$, $i = 1, \cdots, D!$. Clearly, $P$ provides a significant and feasible way to estimate a characteristic probability distribution function for a given time series, which plays a crucial role in extracting statistical complexity information on the underlying dynamical systems \cite{BandtPRL2002,AmigoPTRSA2014}.

The conceptually relatively simple computation of $P$ based on ordinal patterns for the underlying system prompts to another successful application of this framework in terms of statistical complexity measures (SCMs) \cite{kowalskiEntropy2011}. One common way of defining SCMs is taking the product of a normalized Shannon entropy and an associated measure of disequilibrium \cite{LopezPLA1995,kowalskiEntropy2011}, which captures specific organizational properties of structure and patterns in a process. Following the vast amount of previous work on this topic  \cite{rossoPRL2007,LopezPLA1995,kowalskiEntropy2011,Lange2013}, in this work we consider a variant of SCMs based on the concept of Jensen-Shannon divergence, which is defined as $\mathcal{C}_{JS}[P] = \mathcal{H}[P] \cdot  Q_{JS}[P, P_e]$, where $\mathcal{H}[P]$ is the normalized Shannon entropy and $Q_{JS}[P, P_e]$ is the so-called disequilibrium function measuring the distance of the given distribution $P$ to the uniform distribution $P_e$. Accordingly, we define $Q_{JS}[P, P_e] = Q_{0} \cdot JS[P, P_e]$ where $JS[P, P_e]$ is the Jensen-Shannon divergence between $P$ and $P_e$ and $Q_{0}$ is a normalization factor. Furthermore, the complexity--entropy plane $\mathcal{H} \times \mathcal{C}_{JS}$ has been intensively used to distinguish chaotic systems from stochastic ones \cite{rossoPRL2007}. The analysis of complexity--entropy planes has already taken advantage from the ordinal pattern based symbolic encoding of time series and has found various applications \cite{rossoPRL2007,kowalskiEntropy2011}. However, in the complexity--entropy analysis, there are some chaotic maps that could be easily confused with random noise because no clearly separable domains of values are available to differentiate those different types of dynamics as reported in \cite{BorgesAMC2019}. So far, it remains a challenging task to disclose a possible non-monotonic relationship between complexity and entropy \cite{MartinPLA2003}.

It may be noted that in the above discussion on ordinal pattern based time series analysis methods, the transition behavior between patterns has largely remained unconsidered. Recently, OPTN representations have been proposed to precisely capture this transition behavior of the ordinal patterns \cite{MichaelChaos2015,KulpChaos2016,zhangSciRep2017,McCullough2017b,BorgesAMC2019}, which opens a broader perspective beyond the standard ordinal symbolic analysis of time series. In this framework, each ordinal pattern is considered as a vertex in a graph, and a directed and weighted edge connecting two patterns in the graph is established according to the corresponding transition frequency (i.e., the probability with which a given permutation is followed by a certain other one). One of the advantages of the resulting OPTNs is that we can obtain a pronounced distinction between different types of systems based on short time series only \cite{MichaelChaos2015,BorgesAMC2019}.

In this work, we draw upon the recent success of both, complexity--entropy plain analysis and OPTNs, to generalize the traditional SCMs by incorporating pattern transition probabilities. The advantages of this approach are as follows: (i) All SCMs successfully show clear distinction between different types of dynamics, which is demonstrated for time series from numerical models and experimental data. (ii) The generalized SCMs help to capture a consistent relationship between SCMs and chaoticity. (iii) All SCMs are sensitive to dynamical transitions along a bifurcation scenario, including period doubling, band merging, inner and outer crises.

The remainder of this paper is organized as follows: In Section~\ref{sec:methods}, we introduce the methodology employed in our work. We start by discussing two slightly different ways to obtain the transition matrix between ordinal patterns, which provides the basis for defining the corresponding OPTNs (Section~\ref{sec:OPW}). Based on this matrix, we propose in Section~\ref{sec:SCM} to compute SCMs from three different perspectives, including static pattern frequency properties, dynamic pattern transition properties, and a combination of both. In Section~\ref{sec:results1}, we then proceed with describing some selected results for the logistic map as as paradigmatic model system exhibiting various types of periodic and chaotic solutions. We first discuss in some detail one example of a periodic orbit to emphasize the important effects of embedding parameters in Section~\ref{sec:embeddings}. The practical usefulness of SCMs will be further illustrated by studying four exemplary time series from different typical dynamical regimes (Section~\ref{sec:four}). In Section~\ref{sec:plane}, the complexity--entropy planes associated with each SCM will be discussed while the control parameter of the logistic map is systematically increased. In order to characterize the dynamical transitions in this model, we further study the behavior of the SCM values in dependence on the control parameter in Section~\ref{sec:transi} and show that similar results can also be obtained for time-continuous dynamical systems like the R\"ossler oscillator in Section~\ref{sec:cont}. Finally, we show that SCMs successfully distinguish different dynamical regimes in experimental time series from flow data and human ECG measurements in Section~\ref{sec:time}. Some concluding remarks and discussions will be provided in Section~\ref{sec:con}.

\section{Methods} \label{sec:methods}

\subsection{Ordinal pattern transition matrix} \label{sec:OPW}

Most recent studies employing the concept of permutation entropy $\mathcal{S}_O$ have defined this measure by focusing exclusively on the frequencies of ordinal patterns, which however disregards the transition behavior between subsequent patterns. To emphasize the transition frequencies between any pair of patterns, measures of transitional complexity have been further proposed in Refs.~\cite{zhangSciRep2017,McCullough2017b} to quantify the properties of the associated OPTNs. To this end, we associate each directed link in an OPTN with its transition frequency $w_{ij} = p({\pi_i \to \pi_j})$. Calculating the transition frequencies for each pair of patterns, we obtain a weighted directed network characterized by a weighted adjacency matrix $\mathbf{W} = \{ w_{ij} \}, i, j \in [1, D!]$.

Based on this transition matrix $\mathbf{W}$ of an OPTN, we note that there two slightly different ways to introduce normalization factors. The first option is to use a global normalization to ensure $\sum_{i,j}^{D!} w_{ij} = 1$ \cite{zhangSciRep2017}, which results in a globally normalized transition entropy as discussed below. The second option has been proposed by McCullough {\textit{et al}}~\cite{McCullough2017b} who first considered the local out-link transition frequency from pattern $\pi_i$ to $\pi_j$
\begin{equation} \label{eq:localTp}
p_{\pi_i \to \pi_j} = \left \{ \begin{aligned}
				& 0,  \;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\; \text{if} \;\;\; \pi_i = \pi_j \\
				& \frac{w_{ij}}{\sum_{j, j \neq i} w_{ij}}, \;\;\;\;\;\; \text{if} \; \pi_i \neq \pi_j.\\
				\end{aligned}
				\right.
\end{equation}
\noindent
Note that the transition frequency of Eq.~\eqref{eq:localTp} is pattern (row-) wise normalized. This case will be referred to as the node-wise out-link normalized transition matrix $\mathbf{W}$ in the following. In this way, it easily captures the heterogeneous behavior of both (static) occurrence frequencies of different patterns and (dynamic) transition frequencies between patterns.

In addition to the proper normalization, we need to consider self-loops that may affect the numerical estimation of $\mathbf{W}$ especially for time series from continuous systems with a given sampling rate. For example, it has been demonstrated in Ref.~\cite{zhangSciRep2017} that there are about $99\%$ self-loops in the time series of R\"ossler system when integrated with a step size $h = 0.01$, while the about $1\%$ non-self-loops are hidden in $\mathbf{W}$. From this example, one may easily see that self-loops are related to the serial correlation of the time series under study \cite{BorgesAMC2019}. On the other hand, neglecting self-loops in $\mathbf{W}$ emphasizes the transition behavior between different patterns, which has therefore been adopted in most existing studies of complex networks for computational simplicity and theoretical concerns \cite{CostaADPhy2007}. Accordingly, we also remove self-loops in the present study before computing the weighted matrix $\mathbf{W}$ while acknowledging that in other cases of stochastic processes and discrete systems, self-loops have been included in the analysis showing some interesting results \cite{BorgesAMC2019}. Arguably, whether to remove or to consider self-loops depends on the particular process under study. In any of the applications discussed in the rest of this paper, we have ensured that our results do not change qualitatively when considering the effects of self-loops.

\subsection{Statistical complexity measures} \label{sec:SCM}

In the following, we will briefly review the existing ordinal pattern frequency based SCM before proposing two novel ways to define SCMs, which explicitly consider transition frequencies of the resulting OPTN representations for time series.

\subsubsection{SCM based on the permutation entropy of pattern frequencies}

The permutation entropy of a time series is the Shannon entropy of the distribution of patterns $P = \{p_{\pi_1}, p_{\pi_2},\ldots,p_{\pi_i} \}$, which is defined as
\begin{equation}
\mathcal{S}_{O}[P]= - \sum_{i=1}^{D!} p_{\pi_i} \log p_{\pi_i}. 
\end{equation}
The normalized permutation entropy $\mathcal{H}_O$ hence follows as 
\begin{equation} \label{eq:Ho}
\mathcal{H}_{O}[P] = \mathcal{S}_{O}[P] / \mathcal{S}_{O, max}, 
\end{equation}
where $\mathcal{S}_{O, max} = \log D!$ corresponds to the case of a uniform distribution of the same number of patterns, i.e., $P_e = \{1/D!, \ldots, 1/D!\}$.

In the next step, we use some appropriate statistical concept to measure the distance of the given empirical distribution $P$ to the uniform distribution $P_e$, which is the essential idea behind the class of SCMs considered in this work. Several different distance functions have been considered in the literature, including the Euclidean norm, Wooters' distance, Kullback-Leibler relative entropy and Jensen-Shannon divergence \cite{kowalskiEntropy2011}. In this work, the distance between the distributions $P$ and $P_e$ is characterized by the Jensen-Shannon divergence $Q_{JS}[P] = Q_0 \cdot JS[P, P_e]$ with 
\begin{equation}
JS[P, P_e] = \mathcal{S}_O[(P+P_e)/2] - \frac{1}{2}\mathcal{S}_O[P] - \frac{1}{2}\mathcal{S}_O[P_e], 
\end{equation}
where $Q_0$ is a normalization factor with equals the inverse of the maximal Jensen-Shannon divergence $JS[P, P_e]$, which ensures that $Q_{JS}[P] \in [0, 1]$. This maximal distance is obtained when the distribution $P$ has probability one for one pattern and zero for all other patterns. The general expression for $Q_0$ reads  
\begin{equation} \label{eq:Q0}
Q_0 = \frac{-2}{\left(\frac{N+1}{N}\right) \log (N+1) -2 \log (2N) + \log (N)},
\end{equation}
where $N = D!$ is the total number of possible patterns. For large $D$, we have $Q_0 \approx 1 /  \log 2$.

The function $Q_{JS}[P]$ is different from zero if the system has a kind of preference of certain patterns among the accessible ones, reflecting determinism of the system. Finally, the SCM based on permutation entropy of node (pattern) frequencies is defined as 
\begin{equation}
\mathcal{C}_{O}[P] = Q_{JS}[P] \cdot \mathcal{H}_{O}[P].
\end{equation}
This quantity is static in the sense that it quantifies the amount of information stored by ordinal frequencies in the system and its disequilibrium of the observed parts of its accessible states in comparison to a uniform distribution \cite{LopezPLA1995}.

\subsubsection{SCM based on the transition entropy of the globally normalized weight matrix}

In this case, the sum of all entries of the transition matrix $\mathbf{W}$ is 1, and the Shannon entropy of transition frequencies between any pair of ordinal patterns is therefore 
\begin{equation}
\begin{split}
\mathcal{S}_{W}[W] &= - \sum_{i=1}^{D!} \sum_{j = 1; j \neq i}^{D!} w_{ij} \log w_{ij}, \\
& = - \sum_{i=1}^{D!} \sum_{j =1; j \neq i}^{D!} p_{\pi_i \to \pi_j} \log p_{\pi_i \to \pi_j}. 
\end{split}
\end{equation}
The normalized entropy is hence defined 
\begin{equation}
\mathcal{H}_{W}[W] = \mathcal{S}_{W}[W] / \mathcal{S}_{W, max}, 
\end{equation}
where $\mathcal{S}_{W, max} = \log D! (D! - 1)$ as any self transitions are excluded. Furthermore, there are $D! (D! - 1)$ possible pattern transitions in the uniformly distribution function $P_e$. Thus, in a full analogy with the node-based SCM, the Jensen-Shannon distance is computed as $Q_{JS}[W] = Q_0 \cdot JS[W, P_e]$ with
\begin{equation}
JS [W, P_e] = \mathcal{S}_{W}[(W + P_e)/2] - \frac{1}{2}\mathcal{S}_{W}[W] - \frac{1}{2}\mathcal{S}_{W}[P_e]. 
\end{equation}
We note that we have the same expression for the normalization factor $Q_0$ as above while replacing $N = D! (D! - 1)$ in Eq.~\eqref{eq:Q0}. This further yields the SCM 
\begin{equation}
\mathcal{C}_{W}[W] = Q_{JS}[W] \cdot \mathcal{H}_{W}[W].
\end{equation}
This quantity is dynamic in the sense that it considers the state transition information stored in the system.

\subsubsection{SCM based on the transition entropy of the node-wise out-edge normalized weighting matrix}
